---
title: "Population vs CO₂: analysis"
author: "Niyonshuti Patrick | Student ID: 101420"
date: "2025-10-10"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Force UTF-8 where possible (Windows/mac/Linux fallbacks)
options(encoding = "UTF-8")
for (loc in c("en_US.UTF-8","English_United States.utf8","en_GB.UTF-8","C.UTF-8")) {
  try(suppressWarnings(Sys.setlocale("LC_CTYPE", loc)), silent = TRUE)
}

# Helper: make data frames UTF-8 and column-names ASCII-safe
utf8ify_df <- function(df) {
  stopifnot(is.data.frame(df))
  # column names → UTF-8
  nm <- iconv(names(df), from = "", to = "UTF-8", sub = "")
  # transliterate to ASCII (avoids glyphs like ²)
  if (requireNamespace("stringi", quietly = TRUE)) {
    nm <- stringi::stri_trans_general(nm, "Any-Latin; Latin-ASCII")
  }
  # clean to safe names
  nm <- gsub("[^A-Za-z0-9]+", "_", nm)
  nm <- gsub("_+", "_", nm)
  nm <- sub("^_|_$", "", nm)
  # friendly specific fix
  nm <- sub("Area_km2?$", "Area_km2", nm, perl = TRUE)
  names(df) <- nm

  # character columns → UTF-8
  chr <- vapply(df, is.character, TRUE)
  if (any(chr)) {
    df[chr] <- lapply(df[chr], function(x) iconv(x, from = "", to = "UTF-8", sub = ""))
  }
  df
}

```


## Loading of the Datasets
```{r}
world_population <- read.csv("C:/Old Documents/BIG DATA/R-Programming/MID TERM Project/world_population.csv")
View(world_population)
CO2_emission <- read.csv("C:/Old Documents/BIG DATA/R-Programming/MID TERM Project/CO2_emission.csv")
View(CO2_emission)


```
```{r}
## Exploratory data analysis
variable.names(world_population)
```


```{r}
head(world_population,5)
```
```{r}
tail(world_population,10)
```
```{r}
str(world_population)
```
```{r}
dim(world_population)
```
```{r}
```


```{r}
duplicated(world_population)
```
```{r}
sum(duplicated(world_population))
View(sum(duplicated(world_population)))

```
```{r}
colSums(is.na(world_population))
```

## Using a box plot to check if there are any outliers in the quantitative variables

```{r}
# Select only numeric columns
library(ggplot2)
library(reshape2)

# Select only numeric columns
numeric_data <- world_population[sapply(world_population, is.numeric)]

# Normalize numeric columns (0–1 scaling)
numeric_data_norm <- as.data.frame(lapply(numeric_data, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))

# Convert to long format for ggplot
melted_data <- melt(numeric_data_norm, variable.name = "Variable", value.name = "Value")

# Create the boxplot
ggplot(melted_data, aes(x = Variable, y = Value)) +
  geom_boxplot(fill = "darkgreen", color = "green") +
  theme_minimal() +
  labs (
    title = " Boxplot of World Population Variables",
    x = "Variables",
    y = "Normalized Values (0–1)"
  )
```



```{r}
#Function to remove outliers using IQR rule
remove_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  x[x < (Q1 - 1.5 * IQR) | x > (Q3 + 1.5 * IQR)] <- NA  # Replace outliers with NA
  return(x)
}
numeric_data <- world_population[sapply(world_population, is.numeric)]

# Remove outliers in each numeric column
numeric_no_outliers <- as.data.frame(lapply(numeric_data, remove_outliers))
library(reshape2)
library(ggplot2)

# Normalize again after removing outliers
numeric_no_outliers_norm <- as.data.frame(lapply(numeric_no_outliers, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))

# Melt for ggplot
melted_data_no_outliers <- melt(numeric_no_outliers_norm, variable.name = "Variable", value.name = "Value")

# Plot
ggplot(melted_data_no_outliers, aes(x = Variable, y = Value)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  theme_minimal() +
  labs(
    title = "Boxplot of World Population Variables (Outliers Removed)",
    x = "Variables",
    y = "Normalized Values (0–1)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Creating a new variable from world population dataset
```{r}
library(dplyr)
#library(tidyverse)
new_var<-select(world_population,Country.Territory,X2022.Population,Growth.Rate)
head(new_var)

```

```{r}
exponential_growth <- function(P, r, t) {
  Population_2030 <- P * exp(r * t)
  return(Population_2030)
}

population_data <- new_var %>%
   mutate(Population_2030 = X2022.Population * exp((Growth.Rate/100) * 8))
 
head(population_data,10)
```
## 4.4 Value extraction and plot
```{r}
top10_countries<-new_var%>%
  arrange(desc(X2022.Population))
head(top10_countries,10)
top10 <-head(top10_countries,10)
top10
```
```{r}
# Reorder Country.Territory by descending population
top10$Country.Territory <- factor(
  top10$Country.Territory,
  levels = top10$Country.Territory[order(top10$X2022.Population, decreasing = FALSE)]
)

# Plot with flipped coordinates
ggplot(top10, aes(x = Country.Territory, y = X2022.Population)) +
  geom_bar(stat = "identity", fill = "red") +
  coord_flip() +  # horizontal bars
  labs(
    title = "Top 10 Most Populous U.S. Counties (2022)",
    x = "County",
    y = "2022 Population"
  ) +
  theme_minimal(base_size = 14)

```
```{r}
world_pop1990_2022<-world_population[,c("Country.Territory","X1990.Population","X2000.Population","X2010.Population","X2015.Population","X2020.Population","X2022.Population")]
world_pop1990_2022

```
```{r}

top10_pop1990_2022 <- world_pop1990_2022[world_pop1990_2022$Country %in% c("Mexico", "Russia", "Bangladesh", "Brazil", "Nigeria","Pakistan","China","India","United States","Indonesia" ), ]
top10_pop1990_2022

```
```{r}
# library
library(tidyverse)

# Our Data frame
data <- top10_pop1990_2022

# Convert from wide to long format
data_long <- data %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(gsub("X|\\.Population", "", Year)),
         Population_Millions = Population / 1e6)  # convert to millions

# Plot population trend
ggplot(data_long, aes(x = Year, y = Population_Millions, color = Country.Territory)) +
  geom_line(size = 1.4) +
  geom_point(size = 2) +
  labs(
    title = "Population Growth Trend (1990–2022)",
    x = "Year",
    y = "Population (Millions)",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.6, size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )



```
## 4.3 CO2 Emissions Trend for Top 10 Countries
```{r}
CO2_emission
top10_CO2_emission <- CO2_emission[CO2_emission$Country.Name %in% c("Mexico", "Bangladesh", "Brazil", "Nigeria","Pakistan","China","India","United States","Indonesia","Russian Federation" ), ]
top10_CO2_emission

```
```{r}

Top10_Co2_emission_years<-top10_CO2_emission[,c(1,5:34)]
Top10_Co2_emission_years

```
```{r}

# Load required libraries
library(tidyverse)  # includes ggplot2, dplyr, tidyr, etc.

data <- Top10_Co2_emission_years  

# Reshape data from wide to long
data_long <- data %>%
  tidyr::pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "CO2_Emissions"
  ) %>%
  mutate(Year = as.numeric(sub("X", "", Year)))

# Plot
ggplot(data_long, aes(x = Year, y = CO2_Emissions, color = Country.Name)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "CO₂ Emissions Over Time by Country",
    x = "Year",
    y = "CO₂ Emissions",
    color = "Country"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

```
## 4.5 Correlation Analysis
```{r}
#install.packages("corrplot")
library(dplyr); library(corrplot)

corr_vars <- world_population %>%
  select(Area..km.., Density..per.km.., Growth.Rate, World.Population.Percentage) %>%
  mutate(across(everything(), as.numeric)) %>%    # ensure numeric
  na.omit()

cor_matrix <- cor(corr_vars, use = "pairwise.complete.obs")
corrplot(cor_matrix, method = "color", addCoef.col = "black")

```
2. Merge the World Population Dataset and CO2 Emissions Around the World
on country name . After extracting, 2022 population and 2019 CO2 emissions.Using an appropriate methods(visualization, statistical analysis of nu merical values,regression,...), determine the pattern of relationships between
population number and CO2 emissions.

```{r}
# Merge datasets on country name
merged_data <- world_population %>%
  inner_join(CO2_emission, by = c("Country.Territory" = "Country.Name")) 
merged_data
```

```{r}
extract <- merged_data[,c("Country.Territory","X2022.Population","X2019")]
extract
```

```{r}

colSums(is.na(extract))
extract_clnd <- na.omit(extract)
extract_clnd

```

```{r}
# Statistical correlation
correlation_value <- cor(extract_clnd$X2022.Population, extract_clnd$X2019, method = "pearson")
print(paste("Correlation between 2022 Population and 2019 CO2 emissions:", round(correlation_value, 3)))
```
The Result show no clear relationship between pop and CO2 emmision


```{r}
# Simple regression model
model <- lm(X2019 ~ X2022.Population, data = extract_clnd)
summary(model)
```
Since p-value: 0.9163 the relationship is not statistically significant

```{r}
# Visualization - Scatter plot with regression line
ggplot(extract_clnd, aes(x = X2022.Population, y = X2019)) +
  geom_point(color = "darkblue", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Relationship Between 2022 Population and 2019 CO2 Emissions",
    x = "Population in 2022",
    y = "CO2 Emissions in 2019",
    caption = "Data source: World Population & CO2 Emissions datasets"
  ) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma)

```
```{r}
# install.packages("ggplot2"); install.packages("dplyr")  # if needed
library(ggplot2)
library(dplyr)

df <- extract_clnd %>%
  filter(X2022.Population > 0, X2019 > 0) %>%           # avoid log(<=0)
  mutate(
    log_pop = log10(X2022.Population),
    log_co2 = log10(X2019)
  )

ggplot(df, aes(x = log_pop, y = log_co2)) +
  geom_point(color = "darkgreen", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dotted") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Log–Log Relationship Between Population and CO₂ Emissions",
    x = expression(log[10]("Population 2022")),
    y = expression(log[10](CO[2]~" Emissions 2019"))
  )

```

4.6 Comparing CO2 Emissions by continents
```{r}
install.packages("dplyr")       
library(dplyr)              

world_population %>%
  select(Country.Territory, Continent)

# we need to add continent information to CO2 data
# Get continent mapping from population data
world_population[, c("Country.Territory", "Continent")]
continent_mapping <- world_population %>%
  select(Country.Territory, Continent) %>%
  rename(Country = Country.Territory)
```

```{r}
# Merge with CO2 data
co2_with_continent <- CO2_emission %>%
  left_join(continent_mapping, by = c("Country.Name" = "Country")) %>%
  filter(!is.na(Continent))
co2_with_continent$Continent
```

```{r}
# Calculating the total CO2 emissions by continent for 2019
continental_co2 <- co2_with_continent %>%
  mutate(CO2_2019 = as.numeric(X2019.1)) %>%
  filter(!is.na(CO2_2019)) %>%
  group_by(Continent) %>%
  summarise(
    Total_CO2_2019 = sum(CO2_2019, na.rm = TRUE),
    Countries_Count = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(Total_CO2_2019))

```

```{r}
install.packages("ggplot2")
# Bar plot of continental CO2 emissions
ggplot(continental_co2, aes(x = reorder(Continent, Total_CO2_2019), y = Total_CO2_2019)) +
  geom_col(fill = "coral", alpha = 0.8) +
  coord_flip() +
  labs(title = "Total CO2 Emissions by Continent (2019)",
       x = "Continent",
       y = "Total CO2 Emissions (metric tons per capita)") +
  theme_minimal()
```

```{r}
# Calculating the total CO2 emissions by continent for 2019
continental_co2 <- co2_with_continent %>%
  mutate(CO2_2019 = as.numeric(X2019.1)) %>%
  filter(!is.na(CO2_2019)) %>%
  group_by(Continent) %>%
  summarise(
    Total_CO2_2019 = sum(CO2_2019, na.rm = TRUE),
    Countries_Count = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(Total_CO2_2019))
```
```{r}

```

